---
title: "markdown"
author: "kira"
output:  
  html_document:
     toc: true
     toc_float: true
     theme: cerulean
     highlight: tango
---

<style type="text/css">
  body{
  font-size: 14pt;
}
</style>


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = T
                      ,message = F
                      ,warning = F
                      ,fig.width=9, fig.height=8)
library(tidyverse)
options(tigris_use_cache = TRUE)

```

<hr>

# Installing R

R (the programming language) —
https://cran.r-project.org/

Rstudio (the development environment) —
https://www.rstudio.com/products/rstudio/download/

# Overview

R is open source, which means it is free, and many people supply packages, contribute capabilities to it.

Because it is open source and has a pretty good user ecosystem, people have developed a lot of capabilities for R. It can be used for data wrangling, linking to various other programs/APIs, building interactive data-vis apps (with shiny), spatial analysis and mapping, and whatever other stats/data science tasks.

This will show how to do some very basic analysis, wrangling, and visualization with different approaches in R.

# R Syntaxes

A lot of the capabilities are redundant.  For example, in R there are a variety of different syntaxes — different ways of writing the same code. Different syntaxes may have different niche advantages/disadvantages, but you can use based on your preferences. 

For basic manipulation of data, there are three main syntaxes: data.table, base R, and tidyverse. Base R is always part of R, but to use another syntax or package you have to load it. 

Another person’s syntax cheatsheet (which omits data.table for formula syntax) is [here](https://cran.r-project.org/web/packages/data.table/vignettes/datatable-intro.html)

### Base R

```{r base}

# use a common demonstation dataset, bundled in R.
# '<-' is the assignment operator; it defines an object or variable.
cars <- mtcars

# look at first 5 rows of dataset
head(cars)

# (what does that function do?)
?head
# we can also look at documentation for the dataset we loaded
?mtcars

# use base R to check the average mpg
mean(cars$mpg)

# plot wt to mpg
plot(cars$wt, cars$mpg)

# (damn big cars are wasteful)

```


### Tidyverse


```{r tidy}

# let's load tidyverse
library(tidyverse)

# we can make cars a "tibble" or tidy table, and it displays slightly differently:

# base data.frame
cars

# tidy tibble
cars <- tibble(cars)
cars

# a key component of tidyverse syntax is the pipe: %>%
# pipe passes a componenent into a function. So instead of
mean(cars$mpg)
# we can also write
cars$mpg %>% mean()
# or even
cars %>% .$mpg %>% mean()


# Now it's much easier to see the average MPG for different cylinders
cars %>% 
  group_by(cyl) %>% 
  summarise(mean.mpg = mean(mpg),
            count = n())

# we collapsed the cars dataset to a summary table with mean miles per gallon by
# number of cylinders and the row counts for each number of
# cyclinders.

# use ggplot -- tidyverse visualization technique to make a more complex plots

p <- ggplot(cars
       ,aes(x = wt
            ,y = mpg)) +
  geom_smooth( color = 'turquoise'
               ,size = 2) +
  geom_point(aes(color = factor(cyl))) 

p

# graph attributes that should change based on a variable in the dataset will be wrapped in aes(),
#' other qualities like color or size or opacity (alpha), which may be static, should
#' be outside of aes().

# let's improve the labeling somewhat, by adding to the plot object "p" we just
# created. This is something you can do with ggplot.
p +
  scale_x_continuous(name = 'weight') +
  scale_color_discrete(name = 'cylinders') +
  labs(title = 'Miles per gallon as a function of car weight'
     ,subtitle = 'Points represent a car make, colored by engine cylinders. Smoothed line shown.')

# a messy interactive version (if you want to use 1 function from a package without
# loading the entire package, use "::")
plotly::ggplotly(p)

```

### Data.table

I use mostly tidyverse and some base R when I write code. I won’t write anything about data.table because I don’t use it, but here’s a resource [here](https://cran.r-project.org/web/packages/data.table/vignettes/datatable-intro.html)
