---
title: "Spatial R"
author: "kira"
date: "1/12/2021"
output: html_document
---

```{r setup, include=FALSE}
rm(list = ls())
knitr::opts_chunk$set(echo = TRUE,
                      fig.show = "hold",
                      out.width = "45%")
```

## Library overview

Quick glossary of useful spatial libraries--examples will follow below.

`sf`is the key R library for spatial R. You'll likely always want to load whenever doing mapping + spatial analysis

For pulling in census geometries directly from R, `tigris` is extremely useful. `tidycensus` is less straightforward to use, but can pull census estimates from decennial/ACS as well as geometries (although it has some limitations that can be annoying)

For cleaning + manipulating spatial data, `rmapshaper` and `lwgeom` are also both worth noting.

You might see references to `sp` and a host of other libraries on stackoverflow and elsewhere--`sp` provided some similar functionalities but is rapidly getting displaced by `sf`. I recommend trying to avoid older libraries that are falling out of favor. `sf` is really easy and good and it is gaining popularity for a reason.


#### for visualization

For exploratory visualization, I generally use base `plot` (extended by `sf`), or `mapview`.  

For more complex or polished plots, I generally use either `ggplot2` or `leaflet`.

`leaflet` and `mapview` make interactive/zoom-able plots.

## `sf` Overview

`sf` is the library that all the others above will depend. Full documentation for the library is [here](https://r-spatial.github.io/sf/).

Some quick notes:

* All functions are prefixed with `st_`
* `sf` objects are basically data.frames that include a geometry column. The geometry column will represent the spatial features associated with each row.
* The geometry column will be "sticky"--to drop it, you'll have to turn the `sf` into a tibble or other non-spatial data.frame. This is often useful because grouping an `sf` object will cause geometries to union. This can very useful, but it can also be an expensive operation that slows down non-spatial analysis.

## Spatial data sample workflow

Here's a little sample workflow that models most of the above libraries.

First, let's load `sf` and get some geometries for a study area from `tigris`. I'll also use a personal library I made, `xwalks`, to link counties to commuting zones.

```{r pull.data, results = "hide"}
suppressMessages(library(dplyr))
suppressMessages(library(sf))


# let's get some Places in NJ with population attached, using tidycensus
acs.vars <- tidycensus::load_variables(year = 2018, dataset = "acs5")
acs.vars %>% head(3)

nj.pl <- 
  suppressMessages(tidycensus::get_acs(geography = "place",
                               state = "NJ",
                               variables = "B01001_001", # just population
                              geometry = TRUE,
                               year = 2018))

```

```{r parse}
(nj.pl <- nj.pl %>% 
  select(GEOID, place = NAME, population = estimate))

# class of object
class(nj.pl)

# quick plot
plot(nj.pl["population"])

```

## Coordinate reference systems

You might run into some issues caused by the curvature of the earth. As you probably know, maps are projected different ways to represent geography on a flat surface. The projection method is called a coordinate reference system (CRS).

The curvature of the earth is not dealt with in all spatial R libraries. The `lwgeom` package can extend `sf` to allow
"geodetic" functions, which do account for the curved surface when calculating things like distance or area.
See: (https://r-spatial.github.io/lwgeom/reference/geod.html)

You'll also generally have to transform  your spatial datasets to the same CRS to work with them together. Some functions will give a message/warning/error based on whether it expects long/lat or planar geometries.

A CRS can store a lot of information, but common ones can just be represented by a 4+ digit number, called an [EPSG code](https://en.wikipedia.org/wiki/EPSG_Geodetic_Parameter_Dataset). You can use the *espg 4326* for a common long-lat crs.


```{r crs, error = T}
library(lwgeom)

nj.pl <- st_transform(nj.pl, 4326)
st_crs(nj.pl)

# alternate projection -- this puts geometries in meters instead of long/lat
# this projection won't be area-preserving
nj.plm <- st_transform(nj.pl, "+proj=eqdc +lat_0=0 +lon_0=0 +lat_1=33 +lat_2=45 +x_0=0 +y_0=0")
st_crs(nj.plm)

# compare calculated areas
st_area(nj.pl) %>% head()
st_area(nj.plm) %>% head()
(st_area(nj.pl) - st_area(nj.plm)) %>% summary()

# explict call to geod_area, which expects a long/lat projection
st_geod_area(nj.plm)

```


## Spatial joins & filters

These are useful for working across different geography levels.
Let's join places with commuting zones, say.
I'll use a personal library, `xwalks`, to link counties to CZs.

```{r getcos, message=FALSE, results = "hide"}

# get counties from tigris
counties <- 
  tigris::counties(state = "NJ",
                             cb = TRUE,
                             year = 2018)

```

```{r joins, message=FALSE}

# devtools::install_github("https://github.com/kmcd39/xwalks.git")
# use xwalks to add czs
(co2cz <- xwalks::co2cz)

counties <- left_join(counties, co2cz,
                      by = c("GEOID" = "countyfp"))

# clarify column names, so we don't end up with diffferent GEOIDs for Places and
# Counties
counties <- counties %>% 
  select(countyfp = GEOID, county_name = NAME,
         cz, cz_name)

nj.pl <- nj.pl %>% 
  rename(place.id = GEOID)


# First try spatial join (this will cause an issue)
counties <- counties %>% st_transform(4326)
nj <- counties %>% st_intersection(nj.pl)

nj.pl$place.id %>% duplicated() %>% sum()
nj$place.id %>% duplicated() %>% sum()
# we duplicated places!

# this is because very slight overlaps will be included in the interesection. Joins
# can be thrown off by shapes having slightly different resolutions or other issues

# The xwalks pkg has two fcns for dealing with this. (turns smaller geos into points
# to do so)
nj <- xwalks::generate.coterminous.xwalk(nj.pl, counties)
nj$geometry <- nj.pl$geometry # add back in polygon geometries

nj.pl$place.id %>% duplicated() %>% sum()
nj$place.id %>% duplicated() %>% sum()

# places by cz count
nj %>% count(cz_name)

# plots
plot(counties["cz_name"], main = "counties by cz")
plot(nj["cz_name"], main = "places by cz")

```


### Alternatives mapping methods

* ggplot & geom_sf

```{r map, message=FALSE}

nj.czs <- counties %>% group_by(cz, cz_name) %>% summarise(., do_union = T)

library(ggplot2)
ggplot() +
  geom_sf(data = nj, aes(fill = population)) +
  geom_sf(data = st_boundary(nj.czs), color = "red") +
  scale_fill_binned(type = "viridis", n.breaks = 6)

```

* Mapview

```{r map2, message=FALSE}

library(mapview)
mapview(nj.pl, zcol="population")

```
